Follow-up Samples

### 2017-02-22 submission/manuscript_outline_20161024.tex
incorporate Amanda suggestions results and discussion

\textbf{Results:} The bacterial microbiome in pre and post lesion
removal samples for the adenoma group were more similar to each other
than the carcinoma group (P-value \textless{} 0.05). There was no
difference for either the adenoma or carcinoma group in the relative
abundance of any OTU between the pre and post lesion removal (P-value
\textgreater{} 0.05). A model built to classify model built to classify
lesion had an AUC range of 0.811 - 0.866 while a model built to classify
initial versus follow up samples had an AUC range of 0.641 - 0.805.
Follow up samples had a decrease in the positive probability (P-value
\textless{} 0.05). This decrease in positive probability was also
observed for the initial sample model (P-value \textless{} 0.05). The
lesion model used a total of 53 variables while the initial sample model
used a total of 70 variables. A total of 23 OTUs were common to both
models with the majority of these classifying to commensal bacteria
(e.g. \emph{Bacteroides}, \emph{Clostridiales}, \emph{Blautia}, and
\emph{Ruminococcaceae}).

\textbf{Conclusions:} Our data supports the hypothesis that there are
differences in the bacterial microbiome between pre and post lesion
removal samples. With individuals with carcinoma having more drastic
differences to the overall community then those with adenoma. Changes to
commensal bacteria were some of the most important variables for model
classification, suggesting that these bacteria may be central to initial
polyp formation and transition to carcinoma.

@@

Colorectal cancer (CRC) continues to be a leading cause of cancer
related deaths and is currently the third most common cause of cancer
deaths {[}1,2{]}. Over the last few years death due to the disease has
seen a significant decrease, thanks mainly to improvements in screening
{[}1{]}. However, despite this improvement there are still approximately
50,000 deaths from the disease a year {[}2{]}.

Recently, there has been promising work on the bacterial microbiome and
it's ability to complement existing screening methods such as Fecal
Immunoglobulin Test (FIT) or act alone as a screening tool {[}3,4{]}.
There has also been research into how the bacterial microbiome could be
altered directly on tumor tissue itself {[}5{]}. A few studies have also
shown how this bacterial microbiome {[}6{]} or specific members within
it {[}7{]} could be directly involved with the pathogenesis of CRC.
These studies have helped to provide a tantalizing link between the
bacterial microbiome and CRC. Although these studies suggest that the
bacterial microbiome might change after treatment there remains limited
information on the bacterial microbiome before and after surgical
removal of lesion (adenoma or carcinoma) and whether the community
changes at all.

In this study we tested the hypothesis that the bacterial microbiome
changes between pre (initial) and post (follow up) samples removal of
lesion (adenoma or carcinoma). Our analysis included both alpha and beta
diversity analysis along with investigation of individual operational
taxonomic units (OTUs). We also utilized Random Forest to build models
to classify either lesion and normal samples or initial and follow up
samples and then subsequently observe how these models, as well as
specific OTUs within them, performed pre and post lesion removal. We
also investigated the models for similar important OTUs to identify the
crucial OTUs for not only classifying lesion or normalsamples but also
initial and follow up.

@@

\textbf{\emph{Bacterial Community and FIT:}} We first wanted to test
whether there were any broad differences between initial and follow up
samples based on lesion being either adenoma or carcinoma. What we found
was that the bacterial community in those with carcinoma were more
dissimilar (as measured by thetayc) to their initial sample then those
with adenoma (P-value \textless{} 0.001) {[}Figure 1a{]}. We also found
that there were larger changes in fecal blood (measured by FIT) for
those with carcinoma versus adenoma (P-value \textless{} 0.0001)
{[}Figure 1b{]}. The broad shift in bacterial community structure before
and after surgery was visualized using NMDS for both adenoma
{[}Figure1c{]} (PERMANOVA \textgreater{} 0.05) and carcinoma {[}Figure
1d{]} (PERMANOVA \textless{} 0.05). Interestingly, when initial and
follow up samples were compared to each other, regardless of whether
they were adenoma or carcinoma (lesion), there was no significant
overall difference between them (PERMANOVA \textgreater{} 0.05). When
investigating more broad alpha diversity metrics there was no difference
found between initial and follow up samples for lesion, adenoma only, or
carcinoma only for any metric tested {[}Table S1{]}. We also observed
that there was no difference in OTU relative abundance between initial
and follow up samples for lesion, adenoma only, or carcinoma only
{[}Figure S1{]}.

\textbf{\emph{Carcinoma Associated Bacteria:}} Previous literature has
suggested that a number of oral microbes may be important in CRC
pathogenesis {[}8{]}. So we next examined whether there were differences
in previously well described carcinoma associated OTUs. These included
the OTUs that aligned with \emph{Porphyromonas asaccharolytica}
(Otu000202), \emph{Fusobacterium nucleatum} (Otu000442),
\emph{Parvimonas micra} (Otu001273), and \emph{Peptostreptococcus
stomatis} (Otu001682). There was a decrease in relative abundance in
initial and follow up samples for lesion and carcinoma for
\emph{Parvimonas micra} (P-value \textless{} 0.05), and
\emph{Porphyromonas asaccharolytica} (P-value \textless{} 0.05)
{[}Figure 2A{]}. In contrast, there was no difference in relative
abundance in any of these OTUs for those with adenoma {[}Figure 2B{]}.
We also observed that only a small percentage of those with adenoma or
carcinoma were positive or had an appreciable relative abundance of any
of these respective OTUs {[}Figure 2{]}.

\textbf{\emph{The Lesion Model:}} We next wanted to identify if there
were any common bacterial microbiome changes in individuals with adenoma
or carcinoma versus normal control individuals. We investigated this by
creating a model to classify lesion versus normal based on the bacterial
community and FIT measurements. This model had an AUC range of 0.811 -
0.866 after 100 iterations of 20 repeated 10-fold cross validations. The
ROC curve for the final lesion model used was within the observed range
of the 100 different test set AUC iterations {[}Figure 3A{]}. There were
a total of 53 variables that were used in this model {[}Figure 3B{]}.
The FIT measurement for fecal blood resulted in the largest decrease in
MDA while the OTU with the largest MDA was \emph{Lachnospiraceae}
(Otu000015) {[}Figure 3B{]}.

@@

\textbf{\emph{The Initial Sample Model:}} After building a model to
classify based on lesion we wanted to then build a separate model
specifically to be able to classify whether samples were initial samples
based on the bacterial community and FIT measurements. The initial
sample model had an AUC range of 0.641 to 0.805 after 100 iterations of
20 repeated 10-fold cross validations. The test set AUC range for this
model performed better then the training set AUCs. There was a marked
decrease in the ROC curve for the final model used when compared to the
100 test set AUC iterations {[}Figure 4A{]}. There were a total of 70
variables that were used for this model {[}Figure 4B{]}. The variable
that resulted in the largest MDA was \emph{Pseudomonas} (Otu000438)
while FIT measurement for fecal blood resulted in the sixth largest
decrease in MDA {[}Figure 4B{]}.

@@

\textbf{\emph{Treatment Differences:}} After observing these changes in
the bacterial community and positive probability we wanted to assess
whether additional treatments, such as chemotherapy and radiation, could
have an impact on the results that we observed. There was only a
significant difference for change in positive probability for those
treated with chemotherapy for the initial sample model (P-value
\textless{} 0.05). All other variables that were tested showed no
difference based on whether chemotherapy or radiation was used {[}Table
S3{]}.

@@

From our results there were large observed differences in the bacterial
microbiome between pre and post lesion removal samples based on whether
the individual had an adenoma or carcinoma. There were much larger
differences between initial and follow up samples based on the thetayc
distance metric and in fecal blood as measured by FIT for individuals
with carcinoma versus adenoma {[}Figure 1{]}. However, there were no
differences between initial and follow up samples for Shannon Diversity,
observed OTUs, or evenness regardless of whether the individual had an
adenoma or carcinoma {[}Table S1{]}. There was also no differences in
relative abundance of any specific OTU for lesion, adenoma only, or
carcinoma only {[}Figure S1{]}.

@@

We next created a model that incorporated FIT and the bacterial
microbiome to be able to classify lesions (adenoma or carcinoma). What
we found was that not only was the positive probability of a lesion
reduced after lesion removal {[}Figure 3 \& 4{]} but also that the
commonly associated CRC bacteria were not highly represented within this
model. Rather, OTUs that made up the most important variables
overwhelmingly belonged to commensal bacteria. With only a single OTU
from a previously associated cancer bacterium (\emph{Porphyromonas
asaccharolytica}). We followed this up by creating a second model to
classify initial versus follow up samples. We found that this model was
able to accurately classify initial versus follow up samples {[}Figure 3
\& 4{]}. Providing additional information on the importance of commensal
bacteria was that there were a total of 23 OTUs in common to both models
and the vast majority belonged to regular residents of our gut community
{[}Table S2{]}. Although there was a detectable change in the carcinoma
group and this change was towards what would be expected for normal
controls, the follow up samples may not be a completely normal bacterial
microbiome.

For the majority of tests performed there were no differences in the
bacterial microbiome based on whether chemotherapy or radiation was
received {[}Table S3{]}. Within our study there was a significant
difference for the time elapsed in the collection of the follow up
sample between adenoma and carcinoma (uncorrected P-value \textless{}
0.05), with time passed being less for adenoma (253 +/- 41.3 days) then
carcinoma (351 +/- 102 days). These results would indicate that the
findings described were specific to the surgical intervention and that
some of the differences observed between carcinoma and adenoma samples
could be due to differences in collection time between samples for the
two different groups. Specifically, it could confound the observation
that carcinomas changed more than adenomas {[}Figure 1a \& 1d{]}.

This study builds upon previous work from numerous labs that have looked
into the bacterial microbiome as a potential screening tool
{[}3,4,9--11{]} by exploring what happens to the bacterial community
after surgical removal of a lesion. Based on previous work by Arthur, et
al. {[}12{]} it may not be surprising to have E.coli as one of the most
important OTUs and one that was common to both models. Interestingly,
many of the most important OTUs had taxonomic identification for
resident gut microbes. This could suggest that the bacterial community
is one of the first components that could change during the pathogenesis
of disease. These bacterial microbiome changes could be the first step
in allowing more inflammatory bacterium to gain a foothold within the
colon {[}8{]}.

@@

One limitation of our study is that we do not know whether individuals
who were still classified as positive by the lesion model eventually had
a subsequent CRC diagnosis. This information would help to strengthen
the case for our lesion model keeping a number of individuals above the
cutoff threshold even though at follow up they were diagnosed as no
longer having a lesion. Another limitation is that we do not know if
adding modern tests such as the stool DNA test {[}17{]} could help
improve our overall AUC. This study also drew heavily from those with
Caucasian ancestry making it possible that the observations may not be
representative of those with either Asian or African ancestry. Although
our training and test set are relatively large we still run the risk of
over-fitting or having a model that may not be representative of other
populations. We've done our best to safeguard against this by not only
running 10-fold cross validation but also having over 100 different
80/20 splits to try and mimic the type of variation that might be
expected to occur. The time difference in collection of sample between
adenoma and carcinoma could have affected our observations for
differences between individuals with adenoma or carcinoma. This
confounding though would not affect the observations where these
individuals are grouped together.

Interestingly, within figure 4 the initial sample model showed better
test AUC results then the training set AUC. This may have occurred
because the training AUC determined from 20 repeated 10 fold cross
validation removed samples at random and did not take into account that
they were matched samples. Another potential reason is that the model
itself may be over-fit since the total number of samples was not that
large. However, the lesion model did not suffer from these discrepancies
and similar conclusions can be drawn solely from this model alone.
Further independent studies need to be carried out to verify our
findings since not only are we dealing with feces, which could be very
different than the communities present on the actual tissue, but also
are dealing with correlations that may not be representative of the true
pathogenesis of disease.

@@

\textbf{\emph{Study Design and Patient Sampling:}} The sampling and
design were similar to that reported in Baxter, et al {[}3{]}. In brief,
study exclusion involved those who had already undergone surgery,
radiation, or chemotherapy, had colorectal cancer before a baseline
fecal sample could be obtained, had IBD, a known hereditary
non-polyposis colorectal cancer, or familial adenomatous polyposis.
Samples used to build the models for prediction were collected either
prior to a colonoscopy or between 1 - 2 weeks after. The bacterial
microbiome has been shown to normalize within this time period {[}18{]}.
Our follow up data set had a total of 67 individuals that not only had a
sample as described but also a follow up sample between 188 - 546 days
after lesion removal and treatment had been completed. This study was
approved by the University of Michigan Institutional Review Board. All
study participants provided informed consent and the study itself
conformed to the guidelines set out by the Helsinki Declaration.

@@

\hypertarget{ref-baxter_microbiota-based_2016}{}
3. Baxter NT, Ruffin MT, Rogers MAM, Schloss PD. Microbiota-based model
improves the sensitivity of fecal immunochemical test for detecting
colonic lesions. Genome Medicine. 2016;8:37.

\hypertarget{ref-zeller_potential_2014}{}
4. Zeller G, Tap J, Voigt AY, Sunagawa S, Kultima JR, Costea PI, et al.
Potential of fecal microbiota for early-stage detection of colorectal
cancer. Molecular Systems Biology. 2014;10:766.

\hypertarget{ref-dejea_microbiota_2014}{}
5. Dejea CM, Wick EC, Hechenbleikner EM, White JR, Mark Welch JL,
Rossetti BJ, et al. Microbiota organization is a distinct feature of
proximal colorectal cancers. Proceedings of the National Academy of
Sciences of the United States of America. 2014;111:18321--6.

\hypertarget{ref-zackular_manipulation_2016}{}
6. Zackular JP, Baxter NT, Chen GY, Schloss PD. Manipulation of the Gut
Microbiota Reveals Role in Colon Tumorigenesis. mSphere. 2016;1.

\hypertarget{ref-arthur_microbial_2014}{}
7. Arthur JC, Gharaibeh RZ, Mühlbauer M, Perez-Chanona E, Uronis JM,
McCafferty J, et al. Microbial genomic analysis reveals the essential
role of inflammation in bacteria-induced colorectal cancer. Nature
Communications. 2014;5:4724.

>>>>

\textbf{Results:} The bacterial microbiome changed more in response to
lesion removal in carcinoma cases compared to adenoma cases (P-value
\textless{} 0.05). There was no difference for either the adenoma or
carcinoma group in the relative abundance of any OTU between the pre and
post lesion removal (P-value \textgreater{} 0.05). A model built to
classify lesion had an AUC range of 0.811 - 0.866 while a model built to
classify initial versus follow up samples had an AUC range of 0.641 -
0.805. Follow up samples had a decrease in the positive probability of
lesion (P-value \textless{} 0.05) suggesting a movement towards a more
normal bacterial community. The initial sample model had a decrease in
positive probability for the follow up samples tobe an initial sample
(P-value \textless{} 0.05). The lesion model used a total of 53
variables while the initial sample model used a total of 70 variables. A
total of 23 OTUs were common to both models with the majority of these
classifying to commensal bacteria (e.g. \emph{Bacteroides},
\emph{Clostridiales}, \emph{Blautia}, and \emph{Ruminococcaceae}).

\textbf{Conclusions:} Our data supports the hypothesis that there are
changes in the bacterial microbiome following colorectal cancer lesion
removal. Individuals with carcinoma have more drastic differences to the
overall community then those with adenoma. Changes to commensal bacteria
were some of the most important variables for model classification,
suggesting that these bacteria may be central to initial polyp formation
and transition to carcinoma.

@@

Colorectal cancer (CRC) continues to be a leading cause of cancer
related deaths and is currently the third most common cause of cancer
deaths {[}1,2{]}. Over the last few years rate of disease mortality has
seen a significant decrease, thanks mainly to improvements in screening
{[}1{]}. However, despite this improvement there are still approximately
50,000 deaths from the disease per year {[}2{]}.

Over the last few years there has been research into how the bacterial
microbiome could be altered directly on tumor tissue itself {[}3{]}. A
few studies have also shown how this bacterial microbiome {[}4{]} or
specific members within it {[}5{]} could be directly involved with the
pathogenesis of CRC. These studies have helped to provide a tantalizing
link between the bacterial microbiome and CRC. Building off of these
findings, there has been promising work on the bacterial microbiome and
its ability to complement existing screening methods such as Fecal
Immunoglobulin Test (FIT) or act alone as a screening tool {[}6,7{]}.
Although these studies suggest that the bacterial microbiome might
change after treatment there remains limited information on the
bacterial microbiome response to lesion (adenoma or carcinoma) removal.

In this study we tested the hypothesis that the bacterial microbiome
changes between pre (initial) and post (follow up) removal of lesion
(adenoma or carcinoma). We analyzed changes in alpha and beta diversity
as well as specific Operational Taxonomic Units (OTUs) relative
abundance. We utilized Random Forest to build two models. The first
model was built to classify either lesion and normal samples while the
second was built to classify initial and follow up samples. Subsequent
observations on how these models, as well as specific OTUs within them,
performed pre and post lesion removal helped inform us as to whether
initial and follow up samples were changing and whether it was towards a
more normal bacterial microbiome or not. We then investigated the two
models for similar important OTUs to identify the crucial OTUs for not
only classifying lesion or normal samples but also initial and follow
up.

@@

\textbf{\emph{Bacterial Community and FIT:}} Within our 67 person cohort
we first wanted to test whether there were any broad differences between
initial and follow up samples based on lesion being either adenoma (n =
41) or carcinoma (n = 26). We found that the bacterial community in
those with carcinoma were more dissimilar to their initial sample than
those with adenoma (P-value \textless{} 0.001) {[}Figure 1a{]}. We also
found that there were larger changes in fecal blood (measured by FIT)
for those with carcinoma versus adenoma (P-value \textless{} 0.0001)
{[}Figure 1b{]}. The bacterial community structure before and after
surgery was visualized using NMDS for both adenoma {[}Figure1c{]}
(PERMANOVA \textgreater{} 0.05) and carcinoma {[}Figure 1d{]} (PERMANOVA
\textless{} 0.05). Interestingly, when initial and follow up samples
were compared, regardless of whether the lesions were adenoma or
carcinoma, there was no significant overall difference in beta diversity
(PERMANOVA \textgreater{} 0.05). When investigating alpha diversity
metrics there was no difference found between initial and follow up
samples for lesion, adenoma only, or carcinoma only for any metric
tested {[}Table S1{]}. We also observed that there was no difference in
the relative abundance of any OTU between initial and follow up samples
for lesion, adenoma only, or carcinoma only {[}Figure S1{]}.

\textbf{\emph{Carcinoma Associated Bacteria:}} Previous literature has
suggested that a number of oral microbes may be important in CRC
pathogenesis {[}8{]}. So we next examined whether there were changes in
previously well described carcinoma associated OTUs, such as
\emph{Porphyromonas asaccharolytica} (Otu000202), \emph{Fusobacterium
nucleatum} (Otu000442), \emph{Parvimonas micra} (Otu001273), and
\emph{Peptostreptococcus stomatis} (Otu001682). We first observed that
only a small percentage of those with adenoma or carcinoma were positive
or had an appreciable relative abundance for any of these respective
OTUs {[}Figure 2{]}. Despite this, those with carcinoma had a decrease
in relative abundance from initial to follow up for \emph{Parvimonas
micra} (P-value \textless{} 0.05) and \emph{Porphyromonas
asaccharolytica} (P-value \textless{} 0.05) {[}Figure 2A{]}. In
contrast, there was no difference in relative abundance in any of these
OTUs when considering only those with adenoma {[}Figure 2B{]}.

\textbf{\emph{The Lesion Model:}} We next wanted to identify if there
were any common bacterial microbiome changes in individuals with adenoma
or carcinoma versus normal controls. We investigated this by creating a
model to classify samples as lesion versus normal based on the bacterial
community and FIT measurements. This model had an AUC range of 0.811 -
0.866 after 100 iterations of 20 repeated 10-fold cross validations. The
ROC curve for the final lesion model used was within the observed range
of the 100 different test set AUC iterations {[}Figure 3A{]}. There were
a total of 53 variables that were used in this model {[}Figure 3B{]}.
The FIT measurement for fecal blood resulted in the largest decrease in
MDA while the OTU with the largest MDA was \emph{Lachnospiraceae}
(Otu000015) {[}Figure 3B{]}.

@@

\textbf{\emph{The Initial Sample Model:}} After building a model to
classify based on lesion we built a separate model specifically to be
able to classify whether samples were initial samples based on the
bacterial community and FIT measurements. The initial sample model had
an AUC range of 0.641 to 0.805 after 100 iterations of 20 repeated
10-fold cross validations. The test set AUC range for this model
performed better then the training set AUCs. There was a marked decrease
in the ROC curve for the final model used when compared to the 100 test
set AUC iterations {[}Figure 4A{]}. There were a total of 70 variables
that were used for this model {[}Figure 4B{]}. The variable that
resulted in the largest MDA was \emph{Pseudomonas} (Otu000438) while FIT
measurement for fecal blood resulted in the sixth largest decrease in
MDA {[}Figure 4B{]}.

@@

\textbf{\emph{Treatment Differences:}} After observing these changes in
the bacterial community and positive probability we wanted to assess
whether additional treatments, such as chemotherapy and radiation, could
have an impact on the results that we observed. There was only a
significant difference for change in positive probability for those
treated with chemotherapy for the initial sample model (P-value
\textless{} 0.05). This suggests that those treated for chemotherapy my
have had a larger change from the initial sample then those without such
treatment. All other variables that were tested showed no difference
based on whether chemotherapy or radiation was used {[}Table S3{]}.

@@

This study builds upon previous work from numerous labs that have looked
into the bacterial microbiome as a potential screening tool
{[}6,7,9--11{]} by exploring what happens to the bacterial community
after surgical removal of a lesion. Based on previous work by Arthur, et
al. {[}12{]} it may not be surprising to have E.coli as one of the most
important OTUs and one that was common to both models. Interestingly,
many of the most important OTUs had taxonomic identification for
resident gut microbes. This could suggest that the bacterial community
is one of the first components that could change during the pathogenesis
of disease. These bacterial microbiome changes could be the first step
in allowing more inflammatory bacterium to gain a foothold within the
colon {[}8{]}.

From our results there were large observed differences in the bacterial
microbiome in samples before and after lesion removal based on whether
the individual had an adenoma or carcinoma. In individuals with
carcinoma compared to adenoma there were much larger differences between
initial and follow up samples based on the community beta diversity and
in fecal blood as measured by FIT {[}Figure 1{]}. However, there were no
differences between initial and follow up samples for any alpha
diversity metric measured regardless of whether the individual had an
adenoma or carcinoma {[}Table S1{]}. There was also no differences in
relative abundance of any specific OTU for lesion, adenoma only, or
carcinoma only {[}Figure S1{]}.

@@

We next created a model that incorporated FIT and the bacterial
microbiome to be able to classify lesions (adenoma or carcinoma). Based
on this model we found that the follow up samples were closer to normal
then the initial samples due to a decrease in positive probability for
lesion and that the commonly associated CRC bacteria were not highly
represented within this model with the exception of \emph{Porphyromonas
asaccharolytica} {[}Figure 3{]}. Although there was a detectable change
and this change was towards what would be expected for normal controls,
it should be noted that the follow up samples may not be a completely
normal bacterial microbiome.

After creating the lesion mode we then created a second model to
classify initial versus follow up samples. We found that this model was
able to accurately classify initial versus follow up samples suggesting
that regardless of adenoma or carcinoma there are distinct common
changes within the bacterial microbiome that occurs after lesion removal
{[}Figure 4{]}. Both models had OTUs that overwhelmingly belonged to
commensal bacteria. Providing additional information on the importance
of commensal bacteria was that there were a total of 23 OTUs in common
to both models and the vast majority belonged to regular residents of
our gut community {[}Table S2{]}.

Within our study there was a significant difference for the time elapsed
in the collection of the follow up sample between adenoma and carcinoma
(uncorrected P-value \textless{} 0.05), with time passed being less for
adenoma (253 +/- 41.3 days) than carcinoma (351 +/- 102 days). These
results would indicate that the findings described were specific to the
surgical intervention and that some of the differences observed between
carcinoma and adenoma samples could be due to differences in collection
time between samples for the two different groups. Specifically, it
could confound the observation that carcinomas changed more than
adenomas {[}Figure 1a \& 1d{]}. This confounding though would not affect
the observations where these individuals were grouped together.

@@

One limitation of our study is that we do not know whether individuals
who were still classified as positive by the lesion model eventually had
a subsequent CRC diagnosis. This information would help to strengthen
the case for our lesion model keeping a number of individuals above the
cutoff threshold even though at follow up they were diagnosed as no
longer having a lesion. Another limitation is that we do not know if
adding modern tests such as the stool DNA test {[}17{]} could help
improve our overall AUC. This study also drew heavily from those with
Caucasian ancestry making it possible that the observations may not be
representative of those with either Asian or African ancestry. Although
our training and test set are relatively large we still run the risk of
over-fitting or having a model that may not be representative of other
populations. We've done our best to safeguard against this by not only
running 10-fold cross validation but also having over 100 different
80/20 splits to try and mimic the type of variation that might be
expected to occur.

Interestingly, within the initial sample model the test data performed
better than the training data. This may have occurred because the
training AUC determined from 20 repeated 10 fold cross validation
removed samples at random and did not take into account that they were
matched samples. Another potential reason is that the model itself may
be over-fit since the total number of samples was not that large.
However, the lesion model did not suffer from these discrepancies.
Further independent studies need to be carried out to verify our
findings and models since we are dealing with correlations that may not
be truly representative of the pathogenesis of disease.

@@

\textbf{\emph{Study Design and Patient Sampling:}} The sampling and
design were similar to that reported in Baxter, et al {[}6{]}. In brief,
study exclusion involved those who had already undergone surgery,
radiation, or chemotherapy, had colorectal cancer before a baseline
fecal sample could be obtained, had IBD, a known hereditary
non-polyposis colorectal cancer, or familial adenomatous polyposis.
Samples used to build the models for prediction were collected either
prior to a colonoscopy or between 1 - 2 weeks after. The bacterial
microbiome has been shown to normalize within this time period {[}18{]}.
Our follow up data set had a total of 67 individuals that not only had a
sample as described but also a follow up sample between 188 - 546 days
after lesion removal and treatment had been completed. This study was
approved by the University of Michigan Institutional Review Board. All
study participants provided informed consent and the study itself
conformed to the guidelines set out by the Helsinki Declaration.

@@

\hypertarget{ref-dejea_microbiota_2014}{}
3. Dejea CM, Wick EC, Hechenbleikner EM, White JR, Mark Welch JL,
Rossetti BJ, et al. Microbiota organization is a distinct feature of
proximal colorectal cancers. Proceedings of the National Academy of
Sciences of the United States of America. 2014;111:18321--6.

\hypertarget{ref-zackular_manipulation_2016}{}
4. Zackular JP, Baxter NT, Chen GY, Schloss PD. Manipulation of the Gut
Microbiota Reveals Role in Colon Tumorigenesis. mSphere. 2016;1.

\hypertarget{ref-arthur_microbial_2014}{}
5. Arthur JC, Gharaibeh RZ, Mühlbauer M, Perez-Chanona E, Uronis JM,
McCafferty J, et al. Microbial genomic analysis reveals the essential
role of inflammation in bacteria-induced colorectal cancer. Nature
Communications. 2014;5:4724.

\hypertarget{ref-baxter_microbiota-based_2016}{}
6. Baxter NT, Ruffin MT, Rogers MAM, Schloss PD. Microbiota-based model
improves the sensitivity of fecal immunochemical test for detecting
colonic lesions. Genome Medicine. 2016;8:37.

\hypertarget{ref-zeller_potential_2014}{}
7. Zeller G, Tap J, Voigt AY, Sunagawa S, Kultima JR, Costea PI, et al.
Potential of fecal microbiota for early-stage detection of colorectal
cancer. Molecular Systems Biology. 2014;10:766.
